Data Parallel Training of Neural Networks

### 2019-07-31 paper/paper.tex
ADD: main chapter Parallelizing Neural Networks

\section{Parallelizing Neural Networks} % (fold)
\label{sec:parallelizing_neural_networks}

\subsection{Model Parallelism} % (fold)
\label{sub:model_parallelism}
% definition/how does it work
% pro/contra (compared to data-parallelism)

@@

\subsection{Data Parallelism} % (fold)
\label{sub:data_parallelism}
% definition/how does it work in general
% different approaches: synchronous/asynchronous etc.
% pro/contra (compared to model-parallelism)

@@

\subsection{Hybrid Approaches} % (fold)
\label{sub:hybrid_approaches}
% definition/how does it work
% motivation, pro/contra to just one parallelism technique

% subsection hybrid_approaches (end)

>>>>

\section{Parallelizing Neural Networks} % (fold)
\label{sec:parallelizing_neural_networks}
This section gives an overview of the different algorithms used for distributed training of neural networks.
There are two main parallelization approaches for neural networks: model and data parallelism.
Data parallelism can be divided further into synchronous and asynchronous approaches.
Moreover, model and data parallelism can be combined to hybrid approaches to utilize both techniques as demonstrated, for example, by \cite{krizhevsky2014-One-weird-trick,das2016-Distributed-deep-learning}.

\subsection{Model Parallelism} % (fold)
\label{sub:model_parallelism}
% definition/how does it work
% pro/contra (compared to data parallelism)

% Definition/how it works
In model parallelism, the neural network is partitioned into multiple components.
These model components can then be distributed across multiple machines and trained in parallel.
They are linked by the connections between neurons in different components.
During training, the output from one component is communicated to the next along these connections \cite{dean2012-Large-scale-distributed}.
\autoref{fig:model_parallelism} by Dean et al. \cite{dean2012-Large-scale-distributed} gives an example of this process.

\begin{figure}[ht]
\centering
\includegraphics[width=2.5in]{images/model-parallelism}
\caption{
A schematic illustration by Dean et al. \cite{dean2012-Large-scale-distributed} of partitioning a neural network to distribute it for model parallelism.
The connections printed in bold cross between two model parts.
These are the connections, along which communication is required.
Image source: \cite{dean2012-Large-scale-distributed}.
}
\label{fig:model_parallelism}
\end{figure}

Since we communicate neuron outputs, model parallelism generally scales best when the amount of computation per neuron is high \cite{krizhevsky2014-One-weird-trick}.
This is typically the case in fully-connected layers \cite{krizhevsky2014-One-weird-trick}.
However, fully-connected layers generally require more communication, as they produce more connections between the model components than models with more ``local connectivity structures'' \cite{dean2012-Large-scale-distributed}.

% Advantages
By using more machines---and thus more cores and more memory---model parallelism can substantially reduce the training time.
For example, Dean et al. were able to train large models with significant speedups using up to 144 partitions \cite{dean2012-Large-scale-distributed}.
%
Additionally, model parallelism reduces the required memory per machine \cite{dean2012-Large-scale-distributed}.
The model, including all its parameters, is partitioned and distributed over multiple machines.
Thus, each machine only needs to store its assigned part of the model and the corresponding parameters.
%
This can be a crucial advantage as some models are too large to fit in the memory of a single machine.
It is especially beneficial on GPUs since they typically offer significantly less memory than CPUs and transferring data from the CPU can pose a severe bottleneck \cite{dean2012-Large-scale-distributed}.
The available memory is thus often a limiting factor to the model size on GPUs \cite{dean2012-Large-scale-distributed,chilimbi2014-Project-Adam}.
Using model parallelism, this restriction can be mitigated, and larger models can be trained using multiple GPUs.

There are, however, also some challenges to model parallelism:
If the required time differs greatly between the machines, faster machines might need to wait a lot for slower machines, which can decrease the obtainable speedup \cite{dean2012-Large-scale-distributed}.
Additionally, the communication time can introduce a significant overhead, especially if the number of connections between the model components is high.

@@

\subsection{Data Parallelism} % (fold)
\label{sub:data_parallelism}
% definition/how does it work in general
% different approaches: synchronous/asynchronous etc.
% pro/contra (compared to model parallelism)

In data parallelism, each machine has its own copy of the neural network.
Instead of partitioning the model, data parallelism partitions the training data and distributes it across different machines.
Each machine then processes its assigned training instances and computes local gradients.
To keep the model copies consistent over all machines, the gradients are communicated between the processes.
Instead of communicating and aggregating the gradients, it is also possible to first apply local parameter updates, followed by synchronizing these parameters between the processes.
%
Data parallelism is generally more efficient if the computation per weight is high, as the model parameters (or gradients) are what is communicated \cite{krizhevsky2014-One-weird-trick}.
High computation per parameter is very typical for convolutional layers where the kernel weights are reused for all input pixels \cite{krizhevsky2014-One-weird-trick}.
%
An important hyper-parameter for data parallelism is the batch size.
Since the gradients are usually communicated once per batch, we can reduce the required communication by decreasing the number of batches, that is increasing the batch size \cite{krizhevsky2014-One-weird-trick}.
According to Krizhevsky, data parallelism can be ``arbitrarily efficient'' with a large enough batch size \cite{krizhevsky2014-One-weird-trick}.

There are many different algorithms for synchronizing the model parameters and communicating the gradients.
Depending on the communication pattern, they can be divided into synchronous or asynchronous techniques.
Both approaches are described in the following.

\subsubsection{Synchronous Data Parallelism} % (fold)
\label{ssub:synchronous_data_parallelism}
Synchronous approaches---using, for example, synchronous SGD---usually consist of two phases alternated in lockstep \cite{jin2016-How-to-scale}.
In the first phase, each machine processes its assigned training instances and computes a local gradient.
In the second phase, these local gradients are then aggregated using synchronous communication.
Typically, this communication is performed via an MPI all-reduce operation, which collects all local gradients and re-distributes an aggregated gradient to all processes.

This communication pattern introduces a synchronization barrier in each second phase, where all processes have to wait for each other.
As stated by Jin et al., this synchronization barrier actually makes up the majority of the communication overhead on modern networks \cite{jin2016-How-to-scale}.
It also introduces several problems:
Synchronous SGDs are prone to a straggler effect meaning ``the slowest nodes prevent the rest from making progress'' \cite{jin2016-How-to-scale}.
They are also ``not robust in the face of failing processors or nodes'' \cite{jin2016-How-to-scale} since the all-reduce operation cannot complete with any missing processes.
Additionally, synchronous SGDs might under-utilize the compute-resources when much of the training time is spent waiting for other processes \cite{jin2016-How-to-scale}.

Due to the aforementioned disadvantages, synchronous SGD is often believed to be inferior to asynchronous approaches \cite{chen2016-Revisiting-distributed-synchronous-SGD}.
However, some recent developments demonstrate synchronous SGDs outperforming asynchronous algorithms:
For example, Jin et al. find that on more than $32$ nodes synchronous SGDs scales better than their asynchronous Gossiping-SGD \cite{jin2016-How-to-scale}.
Chen et al. demonstrate their improved synchronous SGDs with additional backup workers offers faster convergence with better accuracy than the compared asynchronous training \cite{chen2016-Revisiting-distributed-synchronous-SGD}.

% subsubsection synchronous_data_parallelism (end)

\subsubsection{Asynchronous Data Parallelism} % (fold)
\label{ssub:asynchronous_data_parallelism}

Asynchronous approaches typically try to avoid the synchronization barrier inherent to synchronous SGD.
%
Many asynchronous techniques use a global parameter server approach.
These techniques use two types of processes: a parameter server and multiple worker processes.
The \emph{parameters server} contains the most recent consensus of the model parameters.
Multiple \emph{workers} process the training data in parallel and compute local gradients.
However, instead of communicating directly with other workers, the local gradients are sent to the parameter server, which receives them and updates its model parameters accordingly.
The worker processes periodically fetch the latest parameters from the server to update their copy of the model.
This process is also illustrated in \autoref{fig:parameter_server} by Dean et al. \cite{dean2012-Large-scale-distributed}.

\begin{figure}[ht]
\centering
\includegraphics[width=2.5in]{images/parameter-server.pdf}
\caption{
A schematic illustration of a parameter server by Dean et al. \cite{dean2012-Large-scale-distributed}.
The parameter server receives gradients $\delta w$ from the workers and applies them to its model parameters.
Workers can fetch parameters $w$ from the server to update their local model replicas.
Each worker has its own share of the training data---labeled as data shards in this diagram---which it uses to compute the local gradients.
Image source: \cite{dean2012-Large-scale-distributed}.
}
\label{fig:parameter_server}
\end{figure}

There are different approaches at which times communication takes place.
The simplest approach, as described by Dean et al., is sending a gradient and fetching parameters after every mini-batch---that is, the subset of a batch assigned to this worker \cite{dean2012-Large-scale-distributed}.
They also suggest reducing the commutation by increasing this interval to multiple batches.
The interval for sending gradients and fetching parameters are not required to be equal \cite{dean2012-Large-scale-distributed}.

Since the workers only communicate with the parameter server and never with other workers, they are independent of each other.
There is, thus, no synchronization barrier.
However, asynchronous gradient descent can have its own problems:
Since all workers communicate all-to-one with the parameter server, a communication bottleneck at the parameter server can occur, while other parts of the network might be underutilized \cite{jin2016-How-to-scale}.
Furthermore, the convergence can be worse than for synchronous approaches since the workers usually have a slightly outdated copy of the model \cite{jin2016-How-to-scale,chen2016-Revisiting-distributed-synchronous-SGD}.
This effect is amplified when the communication delay is increased.

Multiple different variations of the classical parameter server approach have been developed, such as Elastic Averaging SGD by Zhang et al. \cite{zhang2015-Elastic-AvgSGD} and Gossiping SGD by Jin et al. \cite{jin2016-How-to-scale}.
There are also approaches eliminating the global parameter server altogether, like Sergeev et al., who utilize ring-all-reduce operations in their Horovod library to overcome the need for a parameter server \cite{sergeev2018horovod}.

% subsubsection asynchronous_data_parallelism (end)

@@
